Preliminary Thesis Defence

### 2014-01-06 south-africa.tex
spelling

\section{Introduction}
\label{sa:intro}
This paper looks at the effect of the gender of pension recipients on the growth of children in the same household.
The study is based on South-African data and the approach is very similar to \textcite{duflo2000child,duflo2003grandmothers},
and orignally based on the work of \textcite{thomas1994like}.
The difference from international standards \parencite[WHO Child Growth Standards]{who2006child} for anthropometrics are computed as z-scores.
Using these standardised metrics, we compare children living in household with pension recipients of different gender.

This study deviates from the Duflo study in several ways.
The core contribution of this paper is the analyis of exogenous change in men's pension eligibility age,
which is the main explanatum.
The pension eligibility age for men was lowered from 65 to 60 between mid 2009 and 1-1-2011.
This brought the pension eligibility age for men at par with women.
There are two reasons why this warants a further look at this topic in this dataset.

Firstly, life expectancy is South Africa is substantialy below the pension eligibility age.
Around the end of the first decade of the century, which is when our data was collected,
the average life expectancy at birth was only slighly above 50 years old.
In the year 1993, when \textcite{duflo2000child,duflo2003grandmothers} are analysed,
the average life expectancy at birth was somewhat higher (slightly above 55 years old).
A further discussion of this can be found is \autoref{sa:results}
However, in both cases, there is a substantial selection bias in the pension recipient base.
Moreover, the fact that men receive pension only at 65, and women at sixty,
causes an even more pronounced selection bias in the male pension recipient base.
This also makes the comparing of the effect of male pension reicipient and female pension recipients,
on the anthropometrics of children in the same household problematic,
since much more attrition will have take place in the male pension base.
The drop in life expectancy also aggravates the issue of attrition,
and the associated selection bias.
However, since pension eligibility becomes equal, for both men and women, at least,
provides us with effects which internally more comparable.
We say more, because the difference in attribition between the male and female pension base in our sample has not entirely been eliminated.
Throughout the evolution of the average life expectancy at birth in South Africa,
female life expectancy has been higher than male life expectancy by about one year.
Bearing in mind that the difference between average life expectancy and pension eligibility is around 9 years,
this additional selection bias effect, should not be underestimated.
Furthermore, is seems likely that a healthy lifestyle, which increases the chance of becoming a pension recipient,
also has an effect on the lifestyle of household members.
In this is so, then our observed attrition will be an actual cause of a selection bias \textbf{effect}.

@@

The impetus for this paper lies in the optimal design of cash transfer schemes such as CCTs and UCTs.
The lack Paretto optimal allocation of resources within households as discussed in i.a.
\textcite{udry1995gender}, \textcite{udry1996gender}, and \textcite{duflo2004intrahousehold},
indicates the necesity of optimal design in such schemes.
Based on this lack of Paretto optimal allocation,
we have to reject the idea of households acting as a unit in an economic sense.
For the design of cash transfer schemes it is therefore necessary to determine the prefered recipient within the household
\footnote{For a good overview see e.g.~\textcite{haddad1997intrahousehold}}.

As mentioned above, we follow \textcite{duflo2000child,duflo2003grandmothers} general design.
Looking at the gender of pension recipients gives a reasonably clean analysis,
because of it's relative exogeneity.
We therefore use these pension receipts as the Right-Hand Side variables, or explanata.
Anthopometrics for children are used, since these capture well, the effects of both malnutrition and disease,
the two most common health impediments that we are addressing.

The South African pension system is an interesting object of study because of its eligibility criteria.
The primary criterium is the age of the recipient.
In adition to this there is a maximum income threshold.
Outside of this, there are very few criteria.
The reletive general applicablility of the program makes that there are few selection bias issues when studying this.
A thorough, though a points somewhat dated, discussion can be found in \textcite{case1998large}.
Although the pension system was intended as a form of poverty relief for the elder population,
it has also become that for the South-African rural population\parencite{tangwe2013impact}.
Average household income in rural area is much lower than in urban areas.
Pension receipt have therefore formed a large share of household income.
Upon the initial expension to include the black population, in 1991,
this was a much as twice the mean monthly income.

The anthropometrics taken in the NIDS are useful for computing z-scores.
We distinguish between Age Based Z-scores (ABZ) and Height-Based Z-scores (HBZ).
We use two types ABZs and two types of HBZs, for a total of four types of z-scores.
This is described in futher detail in \autoref{sa:data:who}.

@@

\subsection{South Africa: National Income Dynamics Survey}
The main source of data is the National Income Dynamics Survey of South Africa \parencite{saldru2008nids, saldru2012nids, saldru2013nids}.
Like the 1993 survey used by \textcite{duflo2000child, duflo2003grandmothers}, this survey is conducted in cooperation with the World Bank.
Unlike the 1993 survey, this survey does not use a random selection household,
rather it collects data on a representative set of appproximately 10,000 South-African households over time.
Currently three `waves' of data are available, these waves date from 2008, 2012, and 2013 respectively.
The primary information types we use are:
\begin{itemize}
  \item child anthropometrics,
  \item child age (in days)
  \item child gender
  \item adult pension recipient status
  \item adult gender
\end{itemize}
In addition to these variables of interest, we include a number of covariates in the analysis, these are:
\begin{itemize}
  \item Household income
  \item Parents education
\end{itemize}

@@

Children's anthropometrics are taken, these are length/height, weight, and waist.
Using these anthropometics and WHO growth standards, z-scores are calculated.

@@

For example, if we measure a height $x$ for a child of age $y$ (in weeks/months),
then we refer the to WHO tables, find the relevant ideal height and standard deviation for a child of age $y$.
We then substract the ideal height ($\mu_y$) from the observed height,
and divide by the standard devation ($\sigma_y$), like so:
\[
z_{xy} = \frac{x - \mu_y}{\sigma_{y}}
\]
These ideal scores are based on a sample of children from different ethnic populations,
in households which observed a healthy lifestyle.
Any health issues, such as malnutrition or disease will affect these metrics,
by causing the child to be shorter or lighter.
However, it is impossible to distinquish between the different causes of an observed slowed growth.

@@

In general, can distinquish between two types of anthropometric z-scores, the age-based z-scores and the height-based z-scores.
Whereby `based' refers to the reference point at which anthropometrics are standardised.

\subsubsection{Age-Based Z-scores}
The Age-Based Z-scores (ABZs) are consistituted by the Height-for-Age Z-score (HAZ) and the Weight-for-Age Z-score (WAZ).
Since these metrics are age-based, they provide information about all past growth issues.
Any past issues such a malnutrition and disease will have impaired growth,
and these effects will still be captured by today's height.
This also applies to the WAZ, as standard weight is a function of the height,
which is in turn a function of the age.

The ABZs are constructed on a weekly basis up to the age of 60 months, and on a monthly basis thereafer.

@@

In these tables the dependent variable used it defined on the top row.
The second row defines the model used (as defined in \autoref{sa:models}).
The other rows represent the independent variables.
Where \textbf{w\_spen\_w} represents the dummy variable for children living in a household with a state pension eligible woman.
The variable \textbf{w\_spen\_m} is the dummy for the child living in the same household as a male state mension recipient.
The policy variable \textbf{elig.men.60} is a dummy which takes the value \textbf{1} for waves 2 and 3.
An interaction term of the later two is also included as \textbf{eli.men.60:w\_spen\_m}.
Lastly, we unclude de covariate \textbf{w\_h\_tinc} which represents total household income.

@@

As \autoref{sa:ta:haz} and \autoref{sa:ta:waz} shows the Height-for-Age and the Weight-for-Age estimations for all three Right-Hand Side variations give similar results.
For all the height-based z-score estmations, we find that the policy variable \textbf{elig.men.60} has a negative coefficient estimate, which is highly significant (where included).
In all cases the estimate has a p-value of less that \textbf{0.01}.
Meaning that (meta-discussion aside) the probabiliy that this coefficient represents a non-existant relation (Type II error) is less than one percent.
In the model 2 specification of the Weight-for-Height dependent variable, the p-value is even less that 0.001.
However, upon the inclusion of the interaction term (model 3) this falls back to below 0.01.

@@

We thus find a negative effect of the treatment on growth metrics.
Furthermore, for one height-based z-score we also find a negative effect of male pension recipients on growth metrics.
Additionaly, it is suprising that we find significant coefficients for \textbf{w\_spen\_m} in one height-based z-score (\textbf{BMIZ}), but not in the other \textbf{WHZ}.

The last result is surprising, in the sense that it is significant for one dependent variable, but not the other.
Especially considering that the coefficient estimates for \textbf{w\_spen\_m} in the \textbf{WHZ} estimations are all similar to each other,
and roughtly half of the estimates of the \textbf{BMIZ} estimations.

@@

Secondly, analysing data around the policy change allows us to employ a Difference-in-Differences estimation (DiD).
Which enables us to draw a causal inference from the treatment effect,
which is onthologically more intersting than a correlation.

We use the South-African National Income Dynamics Survey data
\parencite[in collaboration with The World Bank]{saldru2008nids,saldru2012nids,saldru2013nids}.
This full-panel dataset provides observations from 2008, 2012, and 2013.
The treatment, a policy change which lowers the pension elgilibity age for men from 65 to 60, takes place around 2009.
Since we have only one time period before the treatment (the year 2008), we cannot establish the common trend.
This common trend is used to validate the common trend assumption, which unpins the Difference-in-Differences method.
As discussed in \autoref{sa:results}, we find that not establishing the trend proves problematic for our analysis.
Between the years 2008 and 2012,
we see a substantial drop in the number of children living in the same household as a female pension recipient.
Since female pension recipients are unaffected by the policy change, we have to assume that a further exogenous factor is the cause of this drop.

In our estimations we use the Difference-in-Differences method,
which is operationalised a fixed-effect panel estimation with a time effect.
The contruct three Right-Hand Side (RHS) model.
The first model is a fixed-effect estimation without the treatment dummy.
The second model is a fixed-effects estimation with the treatment dummy (\textbf{elig.men.60}).
the third model is a fixed-effects estimation with the treatment dummy,
as well as an interaction term with the male pension recipient dummy (\textbf{w\_spen\_m}).

@@

Secondly, we find that living with male pension recipient (\textbf{w\_spen\_m}),
is negatively correlated with a the BMI z-scores (\textbf{BMIZ}).
As discussed in \autoref{sa:results}, the Body-Mass-Index Z-score is a convex mapping of the Weight-for-Height z-score.
This can explain the fact that these essentialy similar estimations, give different significance levels.
Considering that the use of the Body Mass Index is far more wide spread than the Weight for Height metric,
we believe that the significance of these coefficients carries more weight.
The fact that these coefficients are negative, indicates that for this short-term z-score,
the prefered recipient would be female household members.

@@

The later result seems to indicate that, at least in the short-run,
the prefered recipient of cash transfers would be a female household member.
As discussed, the selection bias issue in age,
leads to the need for a careful interpretation of the possible external validity of this results.

>>>>

\section{Introduction}
\label{sa:intro}
This paper looks at the effect of the gender of pension recipients on the growth of children in the same household.
The study is based on South-African data and the approach is very similar to \textcite{duflo2000child,duflo2003grandmothers},
and originally based on the work of \textcite{thomas1994like}.
The difference from international standards \parencite[WHO Child Growth Standards]{who2006child} for anthropometrics are computed as z-scores.
Using these standardised metrics, we compare children living in household with pension recipients of different gender.

This study deviates from the Duflo study in several ways.
The core contribution of this paper is the analysis of exogenous change in men's pension eligibility age,
which is the main explanatum.
The pension eligibility age for men was lowered from 65 to 60 between mid 2009 and 1-1-2011.
This brought the pension eligibility age for men at par with women.
There are two reasons why this warrants a further look at this topic in this dataset.

Firstly, life expectancy is South Africa is substantially below the pension eligibility age.
Around the end of the first decade of the century, which is when our data was collected,
the average life expectancy at birth was only slightly above 50 years old.
In the year 1993, when \textcite{duflo2000child,duflo2003grandmothers} are analysed,
the average life expectancy at birth was somewhat higher (slightly above 55 years old).
A further discussion of this can be found is \autoref{sa:results}
However, in both cases, there is a substantial selection bias in the pension recipient base.
Moreover, the fact that men receive pension only at 65, and women at sixty,
causes an even more pronounced selection bias in the male pension recipient base.
This also makes the comparing of the effect of male pension recipient and female pension recipients,
on the anthropometrics of children in the same household problematic,
since much more attrition will have take place in the male pension base.
The drop in life expectancy also aggravates the issue of attrition,
and the associated selection bias.
However, since pension eligibility becomes equal, for both men and women, at least,
provides us with effects which internally more comparable.
We say more, because the difference in attrition between the male and female pension base in our sample has not entirely been eliminated.
Throughout the evolution of the average life expectancy at birth in South Africa,
female life expectancy has been higher than male life expectancy by about one year.
Bearing in mind that the difference between average life expectancy and pension eligibility is around 9 years,
this additional selection bias effect, should not be underestimated.
Furthermore, is seems likely that a healthy lifestyle, which increases the chance of becoming a pension recipient,
also has an effect on the lifestyle of household members.
In this is so, then our observed attrition will be an actual cause of a selection bias \textbf{effect}.

@@

The impetus for this paper lies in the optimal design of cash transfer schemes such as CCTs and UCTs.
The lack Pareto optimal allocation of resources within households as discussed in i.a.
\textcite{udry1995gender}, \textcite{udry1996gender}, and \textcite{duflo2004intrahousehold},
indicates the necessity of optimal design in such schemes.
Based on this lack of Pareto optimal allocation,
we have to reject the idea of households acting as a unit in an economic sense.
For the design of cash transfer schemes it is therefore necessary to determine the preferred recipient within the household
\footnote{For a good overview see e.g.~\textcite{haddad1997intrahousehold}}.

As mentioned above, we follow \textcite{duflo2000child,duflo2003grandmothers} general design.
Looking at the gender of pension recipients gives a reasonably clean analysis,
because of it's relative exogeneity.
We therefore use these pension receipts as the Right-Hand Side variables, or explanata.
Anthropometrics for children are used, since these capture well, the effects of both malnutrition and disease,
the two most common health impediments that we are addressing.

The South African pension system is an interesting object of study because of its eligibility criteria.
The primary criterium is the age of the recipient.
In addition to this there is a maximum income threshold.
Outside of this, there are very few criteria.
The relative general applicability of the program makes that there are few selection bias issues when studying this.
A thorough, though a points somewhat dated, discussion can be found in \textcite{case1998large}.
Although the pension system was intended as a form of poverty relief for the elder population,
it has also become that for the South-African rural population\parencite{tangwe2013impact}.
Average household income in rural area is much lower than in urban areas.
Pension receipt have therefore formed a large share of household income.
Upon the initial expansion to include the black population, in 1991,
this was a much as twice the mean monthly income.

The anthropometrics taken in the NIDS are useful for computing z-scores.
We distinguish between Age Based Z-scores (ABZ) and Height-Based Z-scores (HBZ).
We use two types ABZs and two types of HBZs, for a total of four types of z-scores.
This is described in further detail in \autoref{sa:data:who}.

@@

\subsection{South Africa: National Income Dynamics Survey}
The main source of data is the National Income Dynamics Survey of South Africa \parencite{saldru2008nids, saldru2012nids, saldru2013nids}.
Like the 1993 survey used by \textcite{duflo2000child, duflo2003grandmothers}, this survey is conducted in cooperation with the World Bank.
Unlike the 1993 survey, this survey does not use a random selection household,
rather it collects data on a representative set of approximately 10,000 South-African households over time.
Currently three `waves' of data are available, these waves date from 2008, 2012, and 2013 respectively.
The primary information types we use are:
\begin{itemize}
  \item child anthropometrics,
  \item child age (in days)
  \item child gender
  \item adult pension recipient status
  \item adult gender
\end{itemize}
In addition to these variables of interest, we include a number of covariates in the analysis, these are:
\begin{itemize}
  \item Household income
  \item Parents education
\end{itemize}

@@

Children's anthropometrics are taken, these are length/height, weight, and waist.
Using these anthropometrics and WHO growth standards, z-scores are calculated.

@@

For example, if we measure a height $x$ for a child of age $y$ (in weeks/months),
then we refer the to WHO tables, find the relevant ideal height and standard deviation for a child of age $y$.
We then subtract the ideal height ($\mu_y$) from the observed height,
and divide by the standard deviation ($\sigma_y$), like so:
\[
z_{xy} = \frac{x - \mu_y}{\sigma_{y}}
\]
These ideal scores are based on a sample of children from different ethnic populations,
in households which observed a healthy lifestyle.
Any health issues, such as malnutrition or disease will affect these metrics,
by causing the child to be shorter or lighter.
However, it is impossible to distinguish between the different causes of an observed slowed growth.

@@

In general, can distinguish between two types of anthropometric z-scores, the age-based z-scores and the height-based z-scores.
Whereby `based' refers to the reference point at which anthropometrics are standardised.

\subsubsection{Age-Based Z-scores}
The Age-Based Z-scores (ABZs) are constituted by the Height-for-Age Z-score (HAZ) and the Weight-for-Age Z-score (WAZ).
Since these metrics are age-based, they provide information about all past growth issues.
Any past issues such a malnutrition and disease will have impaired growth,
and these effects will still be captured by today's height.
This also applies to the WAZ, as standard weight is a function of the height,
which is in turn a function of the age.

The ABZs are constructed on a weekly basis up to the age of 60 months, and on a monthly basis thereafter.

@@

In these tables the dependent variable used it defined on the top row.
The second row defines the model used (as defined in \autoref{sa:models}).
The other rows represent the independent variables.
Where \textbf{w\_spen\_w} represents the dummy variable for children living in a household with a state pension eligible woman.
The variable \textbf{w\_spen\_m} is the dummy for the child living in the same household as a male state pension recipient.
The policy variable \textbf{elig.men.60} is a dummy which takes the value \textbf{1} for waves 2 and 3.
An interaction term of the later two is also included as \textbf{eli.men.60:w\_spen\_m}.
Lastly, we include the covariate \textbf{w\_h\_tinc} which represents total household income.

@@

As \autoref{sa:ta:haz} and \autoref{sa:ta:waz} shows the Height-for-Age and the Weight-for-Age estimations for all three Right-Hand Side variations give similar results.
For all the height-based z-score estimations, we find that the policy variable \textbf{elig.men.60} has a negative coefficient estimate, which is highly significant (where included).
In all cases the estimate has a p-value of less that \textbf{0.01}.
Meaning that (meta-discussion aside) the probability that this coefficient represents a non-existent relation (Type II error) is less than one percent.
In the model 2 specification of the Weight-for-Height dependent variable, the p-value is even less that 0.001.
However, upon the inclusion of the interaction term (model 3) this falls back to below 0.01.

@@

We thus find a negative effect of the treatment on growth metrics.
Furthermore, for one height-based z-score we also find a negative effect of male pension recipients on growth metrics.
Additionally, it is surprising that we find significant coefficients for \textbf{w\_spen\_m} in one height-based z-score (\textbf{BMIZ}), but not in the other \textbf{WHZ}.

The last result is surprising, in the sense that it is significant for one dependent variable, but not the other.
Especially considering that the coefficient estimates for \textbf{w\_spen\_m} in the \textbf{WHZ} estimations are all similar to each other,
and roughly half of the estimates of the \textbf{BMIZ} estimations.

@@

Secondly, analysing data around the policy change allows us to employ a Difference-in-Differences estimation (DiD).
Which enables us to draw a causal inference from the treatment effect,
which is ontologically more interesting than a correlation.

We use the South-African National Income Dynamics Survey data
\parencite[in collaboration with The World Bank]{saldru2008nids,saldru2012nids,saldru2013nids}.
This full-panel dataset provides observations from 2008, 2012, and 2013.
The treatment, a policy change which lowers the pension eligibility age for men from 65 to 60, takes place around 2009.
Since we have only one time period before the treatment (the year 2008), we cannot establish the common trend.
This common trend is used to validate the common trend assumption, which unpins the Difference-in-Differences method.
As discussed in \autoref{sa:results}, we find that not establishing the trend proves problematic for our analysis.
Between the years 2008 and 2012,
we see a substantial drop in the number of children living in the same household as a female pension recipient.
Since female pension recipients are unaffected by the policy change, we have to assume that a further exogenous factor is the cause of this drop.

In our estimations we use the Difference-in-Differences method,
which is operationalised a fixed-effect panel estimation with a time effect.
The construct three Right-Hand Side (RHS) model.
The first model is a fixed-effect estimation without the treatment dummy.
The second model is a fixed-effects estimation with the treatment dummy (\textbf{elig.men.60}).
the third model is a fixed-effects estimation with the treatment dummy,
as well as an interaction term with the male pension recipient dummy (\textbf{w\_spen\_m}).

@@

Secondly, we find that living with male pension recipient (\textbf{w\_spen\_m}),
is negatively correlated with a the BMI z-scores (\textbf{BMIZ}).
As discussed in \autoref{sa:results}, the Body-Mass-Index Z-score is a convex mapping of the Weight-for-Height z-score.
This can explain the fact that these essentially similar estimations, give different significance levels.
Considering that the use of the Body Mass Index is far more wide spread than the Weight for Height metric,
we believe that the significance of these coefficients carries more weight.
The fact that these coefficients are negative, indicates that for this short-term z-score,
the preferred recipient would be female household members.

@@

The later result seems to indicate that, at least in the short-run,
the preferred recipient of cash transfers would be a female household member.
As discussed, the selection bias issue in age,
leads to the need for a careful interpretation of the possible external validity of this results.
